Sperm should evolve to make female meiosis fair.

### 2014-05-18 writeup_and_notes/male_dependent_drive.tex
merged figures 1 and 2 to form selfprom.eps. modified legend and in text refs to this fig

In the standard single-locus, biallelic model of female meiotic drive,
the driving allele is transmitted to the egg in heterozygotes with
probability  $d > 1/2$, regardless of sperm genotype \cite[e.g. ][ and see Model 1 in the Appendix for more details]{Ubeda2004}. %The timing of fertilization relative to female meiosis places another constraint on $d$, for example, if fertilization (and therefore, sperm dependent drive) takes place at MII (as in mammals),
%	female drive requires an uneven number of crossovers between the centromere and the drive locus, 
%	so $d$ is bounded to be $<0.83$ \citep[see ][ for discussion]{Buckler1999}.
	%YB - MOVED THIS TO THE APPENDIX
          To depict a case
of a self-promoting meiotic driver,  we modify this standard model such 
	that the driver is only effective when fertilized by a sperm carrying that allele (see Figure \ref{Eggsperm_2_allele_cartoon} and Model 2 in the Supplementary Worksheet). 
We then identify the conditions allowing for the spread of this self-promoting driver, 
	and evaluate whether a driver of this form could generate a sustained conflict favoring the evolution of suppressors. 
We concluded with an analysis of a related model (Model 3) - in which
drivers influence their transmission in females via paternal genotype,
rather than sperm haplotype.

For comparative purposes, we briefly present the standard drive model %as Model 1 in the Appendix, 
%equations \ref{eqn.A1}-\ref{eqn.A3}, 
	%and refer readers to the previous literature on this model 
	\citep[see e.g. ][ for additional results]{Prout1973,Ubeda2004}. 
Assuming the driving allele is deleterious in both sexes, but fully recessive 
	(i.e. the fitness of drive homozygotes equals $w_{BB}=1-s$ and other genotypic fitnesses equal $w_{AA}=w_{AB}=1$), 
	it always invades because,  when rare, it occurs predominantly in heterozygotes and therefore drives without a fitness cost. 
However, when $s$ is large ($s>(2d-1)/(2)$), a driver cannot fix, and
will be maintained as a protected polymorphism \citep{Prout1973}. 
The parameter space where the allele can invade but not fix is shown in white
        in Figure \ref{Invasion_space}A. %, and it can fix in the colored regions. 
When the allele is maintained as a polymorphism, it provides an opportunity for the evolution of
	drive suppressors, corresponding well to empirical examples of
        female meiotic drive \citep[reviewed in ][]{Burt2006}.

In contrast to a traditional driver, which drives but pays effectively
	no fitness cost when rare, 
a self-promoting driver specifically creates low fitness drive homozygotes 
	by uniting driving female gametes with sperm enabling that drive.
        % zygotes when it drives. in heterozygotes. 
%, and therefore there is a drive-associated fitness cost even when rare. 
It must therefore overcome a drive-associated homozygous fitness cost simply to spread when rare. 
The conditions allowing the invasion of a self-promoting driver
 	are consequently far more restrictive than those for a standard meiotic driver.
When rare, a fully recessive, self-promoting driver can only invade when $s$ 
	%, the selection against the drive homozygote, 
	is less than $(2 d - 1)/(4 d)$. 
This analytical approximation, derived from Equation \eqref{deltadriver} assuming Hardy-Weinberg, 
	closely matches results obtained by exact numerical iteration (Figure \ref{Invasion_space}).

When a self-promoting driver does spread, 
	it spends much time at low frequency, 
	because the paucity of complimentary sperm compromises its ability to drive. 
However, once relatively common, it  rapidly achieves fixation due to its
	positive frequency dependent behavior (Inset in Figure \ref{Invasion_space}).  
%This lack of a balanced polymorphism %(Figure \ref{Invasion_space}) 
%	precludes the evolution of an
%	allele that suppresses this form of meiotic drive in females.
The positive frequency dependence of a self-promoting driver can also
	induce bistability in its spread -- some values of $s$ 
	allow the fixation of this driver when common, but 
	preclude its invasion when rare (Equation \ref{ineqA}). 
In this case, the driver will be fixed if it reaches a frequency greater than some threshold 
	(approximated in Equation \ref{bistab_homozy_approx} and  presented exactly in 
	Figure \ref{Bistab_homozyg_cost_fig}) and lost otherwise. 
For most parameters, this threshold is likely too high to be reached by drift, 
	and therefore  the fate of self-promoting drivers is determined
	by the more restrictive invasion criteria rather than the fixation criteria.

@@

\begin{figure}
	\rotatebox{90}{
\includegraphics[width = 0.5
          \textwidth]{Figures/Sperm_egg_cartoon_paper_Fig1.eps}}
\caption{transmission probabilities for alleles through female
  meiosis depend on sperm genotype. 2 allele model. \gc{merge with 2,
    rewrite both captions}}  
\label{Eggsperm_2_allele_cartoon}
\end{figure}

@@

\begin{figure}
\includegraphics[width = 0.8 \textwidth]{Figures/invasion_space_recessive_driver.eps}
\caption{{\bf A.} Invasion analysis for a female meiotic drive allele with
  recessive costs (selection coefficient $s$). First consider a simple female meiotic
  drive system, where the allele is transmitted a fraction $d>1/2$ of
  meioses in female heterozygotes, inspective
  of the allele carried by the fertilizing sperm. 
Such an allele can invade the population over all of the parameter space, and will be maintained as a
  protected polymorphism in the white region, and can fix in the
  population in all of the parameter space colored green or orange. 
The solid back line marks the criteria for the allele to fix $s< (2d-1)/2$.
Now consider an allele that is fairly transmitted through
  female meiosis when fertilized by a sperm carrying a different
  allele, but at a fraction $d>1/2$ when a heterozygote egg is fertilized
  by a sperm carrying the allele (as depicted in Figure
  \ref{Eggsperm_2_allele_cartoon}). Such an allele can only invade in
  the region of parameter space colored orange, and it fixes over all
  of this parameter region. The approximate criteria for this allele
  to fix is given by the dashed line showing $s < (2d- 1)/(4d)$. 
 {\bf B.} In the inset we show trajectories of sperm-dependent female drive
 alleles that can invade the population, each allele has $s =0.1$ against the
 homozygotes and $d =0.6$, $0.7$, $0.8$ shown in yellow, blue, and red. 
 \gc{Likely make this Figure 1B.} } \label{Invasion_space}
\end{figure}

>>>>

In the standard single-locus, biallelic model of female meiotic drive,
the driving allele is transmitted to the egg in heterozygotes with
probability  $d > 1/2$, regardless of sperm genotype \cite[e.g. ][ and see Model 1 in the Appendix for more details]{Ubeda2004}. %The timing of fertilization relative to female meiosis places another constraint on $d$, for example, if fertilization (and therefore, sperm dependent drive) takes place at MII (as in mammals),
%	female drive requires an uneven number of crossovers between the centromere and the drive locus, 
%	so $d$ is bounded to be $<0.83$ \citep[see ][ for discussion]{Buckler1999}.
	%YB - MOVED THIS TO THE APPENDIX
          To depict a case
of a self-promoting meiotic driver,  we modify this standard model such 
	that the driver is only effective when fertilized by a sperm carrying that allele (see Figure \ref{Eggsperm}A and Model 2 in the Supplementary Worksheet). 
We then identify the conditions allowing for the spread of this self-promoting driver, 
	and evaluate whether a driver of this form could generate a sustained conflict favoring the evolution of suppressors. 
We concluded with an analysis of a related model (Model 3) - in which
drivers influence their transmission in females via paternal genotype,
rather than sperm haplotype.

For comparative purposes, we briefly present the standard drive model %as Model 1 in the Appendix, 
%equations \ref{eqn.A1}-\ref{eqn.A3}, 
	%and refer readers to the previous literature on this model 
	\citep[see e.g. ][ for additional results]{Prout1973,Ubeda2004}. 
Assuming the driving allele is deleterious in both sexes, but fully recessive 
	(i.e. the fitness of drive homozygotes equals $w_{BB}=1-s$ and other genotypic fitnesses equal $w_{AA}=w_{AB}=1$), 
	it always invades because,  when rare, it occurs predominantly in heterozygotes and therefore drives without a fitness cost. 
However, when $s$ is large ($s>(2d-1)/(2)$), a driver cannot fix, and
will be maintained as a protected polymorphism \citep{Prout1973}. 
The parameter space where the allele can invade but not fix is shown in white
        in Figure \ref{Eggsperm}B. %, and it can fix in the colored regions. 
When the allele is maintained as a polymorphism, it provides an opportunity for the evolution of
	drive suppressors, corresponding well to empirical examples of
        female meiotic drive \citep[reviewed in ][]{Burt2006}.

In contrast to a traditional driver, which drives but pays effectively
	no fitness cost when rare, 
a self-promoting driver specifically creates low fitness drive homozygotes 
	by uniting driving female gametes with sperm enabling that drive.
        % zygotes when it drives. in heterozygotes. 
%, and therefore there is a drive-associated fitness cost even when rare. 
It must therefore overcome a drive-associated homozygous fitness cost simply to spread when rare. 
The conditions allowing the invasion of a self-promoting driver
 	are consequently far more restrictive than those for a standard meiotic driver.
When rare, a fully recessive, self-promoting driver can only invade when $s$ 
	%, the selection against the drive homozygote, 
	is less than $(2 d - 1)/(4 d)$. 
This analytical approximation, derived from Equation \eqref{deltadriver} assuming Hardy-Weinberg, 
	closely matches results obtained by exact numerical iteration (Figure \ref{Eggsperm}B).

When a self-promoting driver does spread, 
	it spends much time at low frequency, 
	because the paucity of complimentary sperm compromises its ability to drive. 
However, once relatively common, it  rapidly achieves fixation due to its
	positive frequency dependent behavior (Figure \ref{Eggsperm}B.1).  
%This lack of a balanced polymorphism %(Figure \ref{Invasion_space}) 
%	precludes the evolution of an
%	allele that suppresses this form of meiotic drive in females.
The positive frequency dependence of a self-promoting driver can also
	induce bistability in its spread -- some values of $s$ 
	allow the fixation of this driver when common, but 
	preclude its invasion when rare (Equation \ref{ineqA}). 
In this case, the driver will be fixed if it reaches a frequency greater than some threshold 
	(approximated in Equation \ref{bistab_homozy_approx} and  presented exactly in 
	Figure \ref{Bistab_homozyg_cost_fig}) and lost otherwise. 
For most parameters, this threshold is likely too high to be reached by drift, 
	and therefore  the fate of self-promoting drivers is determined
	by the more restrictive invasion criteria rather than the fixation criteria.

@@

\begin{figure}
\includegraphics[width = 0.8 \textwidth]{Figures/selfprom.eps}
\caption{{\bf{A.}} A visual depiction of our model of 'self-promoting' driver. 
	Transmission probabilities for alleles through female
 	 meiosis depend on sperm genotype. 
	 The non-driving $A$-allele, and self-promoting $B$-allele
	 are represented by unfilled and filled circles, respectively. 
	 {\bf{B.}} Evolution of a self-promoter and standard driver. 
	 	Assuming that the fitnesses of drive homozygotes and 
		heterozygotes are $1-s$ and $1$, respectively. 
	 	\emph{Main figure:} Boundary conditions for the invasion 
		and fixation of self-promoting and standard meiotic drivers, 
		with drive coefficient, $d$. Colored regions depict exact results, 
		while lines  represent analytical approximations. 
	 \emph{B1:}  Trajectories of sperm-dependent female drive
		 %alleles that can invade the population, 
		 each allele has $s =0.1$ against the
		 homozygotes.  The drive coefficient is denoted by color. 
 }   
\label{Eggsperm}
\end{figure}

%\begin{figure}
%	\rotatebox{90}{
%\includegraphics[width = 0.5
   %       \textwidth]{Figures/Sperm_egg_cartoon_paper_Fig1.eps}}
%\caption{transmission probabilities for alleles through female
%  meiosis depend on sperm genotype. 2 allele model. \gc{merge with 2,
%    rewrite both captions}}  
%\label{Eggsperm_2_allele_cartoon}
%\end{figure}

@@

%\begin{figure}
%\includegraphics[width = 0.8 \textwidth]{Figures/invasion_space_recessive_driver.eps}
%\caption{{\bf A.} Invasion analysis for a female meiotic drive allele with
%  recessive costs (selection coefficient $s$). First consider a simple female meiotic
%  drive system, where the allele is transmitted a fraction $d>1/2$ of
  %meioses in female heterozygotes, inspective
%  of the allele carried by the fertilizing sperm. 
%Such an allele can invade the population over all of the parameter space, and will be maintained as a
%  protected polymorphism in the white region, and can fix in the
%  population in all of the parameter space colored green or orange. 
%The solid back line marks the criteria for the allele to fix $s< (2d-1)/2$.
%Now consider an allele that is fairly transmitted through
%  female meiosis when fertilized by a sperm carrying a different
%  allele, but at a fraction $d>1/2$ when a heterozygote egg is fertilized
%  by a sperm carrying the allele (as depicted in Figure
%  \ref{Eggsperm_2_allele_cartoon}). Such an allele can only invade in
%  the region of parameter space colored orange, and it fixes over all
 % of this parameter region. The approximate criteria for this allele
%  to fix is given by the dashed line showing $s < (2d- 1)/(4d)$. 
% {\bf B.} In the inset we show trajectories of sperm-dependent female drive
 %alleles that can invade the population, each allele has $s =0.1$ against the
% homozygotes and $d =0.6$, $0.7$, $0.8$ shown in yellow, blue, and red. 
% \gc{Likely make this Figure 1B.} } \label{Invasion_space}
%\end{figure}
